{\pkg}: A Python Package for Integrating Computational Models Across Languages and Scales

### 2019-01-04 papers/lang2018/ms.tex
Added advantages/limitations section

\pdfoutput=1
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage[noend]{algpseudocode}
\usepackage{caption}
\algblockdefx[Try]{Try}{EndTry}%
	{\textbf{try}}%
	{\textbf{end try}}
\algcblockdefx[TryExcept]{Try}{Except}{EndTry}%
	{\textbf{except}}%
	{\textbf{end try}}
\algblockdefx[class]{Class}{EndClass}%
	[2]{\textbf{class} {\sc #1}(#2)}%
	{\textbf{end class}}
\algblockdefx[classmethod]{ClassMethod}{EndClassMethod}%
	[2]{\textbf{method} {\sc #1}(#2)}%
	{\textbf{end method}}

@@

%---------------------------------------------------------------------------------
% ABSTRACT (145 words)
\begin{abstract}
Thousands of computational models have been created within the plant biology community in the past two decades that have the potential to be combined into complex integration networks capable of capturing more complex biological processes than possible with isolated models. However, the technological barriers introduced by differences in language and data formats has slowed this progress. We present \href{https://github.com/cropsinsilico/cis_interface}{\pkg}, a Python package for running integration networks with connections between models across languages and scales. {\pkg} coordinates parallel execution of models in Python, C, C++, and Matlab on Linux, Mac OS, and Windows operating systems, and handles communication in a number of data formats common to computational plant modeling. {\pkg} is designed to be user-friendly and can be accessed at {\tt https://github.com/cropsinsilico/cis\_interface}.
\end{abstract}

@@

%---------------------------------------------------------------------------------
% DOMAIN SPECIFIC LANGUAGE
\subsection{Domain Specific Language (DSL)}\label{SS:dsl}
%
One solution is to agree upon a language. The questions of orchestration, communication, and translation become trivial when models are all written in the same language. Beyond selecting a single programming language, many communities have  developed dedicated domain specific languages (DSLs) for model representation. For example, the Systems Biology Markup Language \citep[SBML][]{Hucka2003} is an XML-based format for representing models of biological processes. Scientists can compose their models using SBML markup and then run their model using software designed to parse SBML files. Other DSLs for biological models include LPy \citep{Boudon2012}, CellML \citep{Cuellar2003}, FLAME \citep{Coakley2012}, and BioPAX \citep{Demir2010}. Dedicated model DSLs improve the reusability of models written in the DSL and have the advantage of often implicitly handling data transformation. However, they provide no help for existing models that are not written in the DSL or those that cannot be expressed within the constraints of the DSL. Furthermore, learning a new DSL, in addition to a programming language, can be daunting.

@@

There are also domain specific frameworks for connecting biological models. For example, OpenAlea \citep{Pradal2015} allows users to compose networks from different components including plant models, analysis tools, and visualization tools. Domain specific frameworks are more intuitive and ultimately more flexible than DSLs in the types of operations they allow models to perform since they have access to the full power of a programming language. However, they are stricter in several respects: 1) the model must be written in (or exposable to), the language of the framework and 2) the model must be written in a way that is aware of the framework and/or the format of models with which it will interact. Like DSLs, frameworks also provided limited support for models written without the framework in mind.

@@

{\pkg} was designed to address the questions from \S\ref{S:background} while being:

@@

To accomplish these objectives, {\pkg} is built upon several well-established open-source software packages along with tools developed explicitly for {\pkg}. {\pkg} currently support models written in Python, Matlab, C, and C++ with 
additional Domain Specific Language (DSL) support for LPy models \citep{Boudon2012}. Support for additional languages is planned for future development (See \S\ref{SS:langsupport}). \todo{Statement on novelty}

%---------------------------------------------------------------------------------
% ORCHESTRATION
\subsection{Orchestration}\label{SS:orchestration}
%
{\pkg} is executed via a command line interface (CLI), {\pkgrun} with specification files (\S\ref{SSS:yaml}) as input. Based on the information contained in the specification files, {\pkg} establishes a network of asynchronous communication channels (\S\ref{SS:communication}), and launches the models on new processes (\S\ref{SS:model_execution}).
%At execution, {\pkg} launches models on new processes based on information contained in a YAML specification file (See \S\ref{SSS:yaml} below) and establishes

@@

%---------------------------------------------------------------------------------
% MODEL EXECUTION
\subsubsection{Model Execution}\label{SS:model_execution}
%
{\pkg} launches each model in an integration on its own process, allowing models to complete independent operations in parallel and complete tasks more quickly. For example, the root and shoot growth models from \S\ref{S:intro} require 2.42\,s and 4.06\,s respectively to run for 100 time steps in isolation. If these two models were manually integrated in serial via the command line, the integration would require a total of 6.48\,s. However, the same integration requires only 4.52\,s using {\pkg}, a speed-up of 1.43. The exact speedup provided by parallel integration using {\pkg} depends on how independent the models are and how equally distributed the work is between the models. If the models are fully independent, the speedup is limited by the time required to execute the slowest model. In the example, this would be the shoot growth model which required almost twice the time to execute as the root model. If the models are dependent on one-another, the models will have to wait for messages to be passed and the speedup will not be as great (see \S\ref{S:speedup} for additional discussion of parallelism and the speedups achievable using {\pkg}). \todo{Fix these numbers for most recent version of the code}.

While every model is executed on a new process, how the model is handled depends 
on the language it is written in. Models written in 
interpreted languages (Python and Matlab) are executed on the command line 
with the interpreter. In the case of Matlab, where a significant amount of time is required to start 
the Matlab interpreter (See \S\ref{SS:results_language}), Matlab shared engines are used 
to execute Matlab models. Matlab shared engines are Matlab instances that other processes can 
submit Matlab code to for execution. While shared engines also require the same amount of time 
to start as a standard instance of the Matlab interpreter, they can be started in advance and then 
reused.

@@

The algorithms for a generic model driver and the master thread can be found in Algorithms \ref{alg:model} \& \ref{alg:master} respectively.
% MODEL DRIVER
\begin{algorithm}[H]
    \caption{Generic model driver class. This algorithm does not represent the actual implementation of any one of the languages supported by {\pkg}. Each of the language-specific drivers will have slightly different methods for compilation (turning the source code into an executable), launching the model process, and monitoring the model's status via output and error codes. However, this algorithm does provided the general behavior pattern upon which every model driver is based.}
    \label{alg:model}
    \begin{algorithmic}[1]
    \Class{ModelDriver}{$name, args, yml$}
    	\State $errors \gets []$
    	\State $executable \gets$ \Call{Compile}{args}
	\State $process \gets$ \Call{ModelProcess}{} on a new process
	\State $thread \gets$ \Call{ModelThread}{} on a new thread
	\item[]
	%
	\ClassMethod{ModelProcess}{}
		\State \Call{Executable}{}
	\EndClassMethod
	\item[]
	%
	\ClassMethod{ModelThread}{}
		\While{\Call{IsAlive}{{\sc ModelProcess}}}
			\If{\Call{HasOutput}{{\sc ModelProcess}}}
				\State \textbf{Print} \Call{GetOutput}{{\sc ModelProcess}}
			\Else
				\State \textbf{Sleep}
			\EndIf
		\EndWhile
		\If{\Call{HasErrors}{{\sc ModelProcess}}}
			\State $errors \gets$ \Call{GetErrors}{{\sc ModelProcess}}
		\EndIf
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Terminate}{}
		\State \Call{Kill}{{\sc ModelProcess}}
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Stop}{}
		\While{not $errors$}
			\State \textbf{Sleep}
		\EndWhile
		\State \Call{Terminate}{}
	\EndClassMethod
	\item[]
    \EndClass
    \end{algorithmic}
\end{algorithm}

%
% MASTER THREAD
%\begin{center}
\begin{algorithm}[H]
	\caption{Basic steps executed on the master thread to create, start, and monitor model \& connection drivers. This pseudocode is not an exhaustive reproduction of the entire implementation, but represents the broad strokes needed to understand the purpose and behavior of the master thread. The algorithm for the {\sc ModelDriver} and {\sc ConnectionDriver} can be found in Algorithms \ref{alg:model} \& \ref{alg:conn} respectively.}
	\label{alg:master}
	\begin{algorithmic}[1]
	\State $models$ $\gets$ \textbf{Parse} YAML files
	\State $inputChannels \gets Map()$
	\State $outputChannels \gets Map()$
	% CREATE DRIVERS
	\item[] \Comment Create Drivers
	\ForAll{$m in models$}
		\ForAll{$i in m.inputs$}
			\If{not $i.isFile$}
				\State $inputAddress \gets i.address$
			\Else
				\State $inputAddress \gets `generate'$
			\EndIf
			\State $i.driver \gets$ \Call{ConnectionDriver}{$inputAddress, `generate', i$}
			\State $inputChannels[i.name] = i$
		\EndFor
	\EndFor
	\ForAll{$m in models$}
	\algstore{alg:master}
	\end{algorithmic}
\end{algorithm}
% PAGE BREAK
\begin{algorithm}[H]
	\caption{Continuation of Algorithm \ref{alg:master}}
	\begin{algorithmic}[1]
	\algrestore{alg:master}
		\ForAll{$o \in m.outputs$}
			\If{$o.isFile$}
				\State $outputAddress \gets o.address$
			\Else
				\If{$o.args \in inpurChannels$}
					\State $outputAddress \gets inputChannels[o.args]$
				\Else
					\State \Call{Error}{}
				\EndIf
			\EndIf
			\State $o.driver \gets$ \Call{ConnectionDriver}{$`generate', outputAddress, o$}
			\State $outputChannels[o.name] = o$
		\EndFor
	\EndFor
	\ForAll{$m in models$}
		\State $m.driver \gets$ \Call{ModelDriver}{$m.name, m.args, m$}
	\EndFor
	% START DRIVERS
	\item[] \Comment Start Drivers	
	\ForAll{$m in models$}
		\ForAll{$i in m.inputs$}
			\State $i.driver.$\Call{Start}{}
		\EndFor
	\EndFor
	\ForAll{$m in models$}
		\ForAll{$o \in m.outputs$}
			\State $o.driver.$\Call{Start}{}
		\EndFor
	\EndFor
	\ForAll{$m in models$}
		\State $m.driver.$\Call{Start}{}
	\EndFor
	% MONITOR DRIVERS
	\item[] \Comment Monitor Drivers
	\State $errorFlag \gets false$
	\While{not $errorFlag$}
		\ForAll{$m in models$}
			\If{$m.driver.errors$}
				\State $errorFlag \gets true$
				\State \textbf{break}
			\EndIf
		\EndFor
	\EndWhile
	% STOP DRIVER
	\item[] \Comment Stop Drivers
	\If{$errorFlag$}
		\State $allErrors \gets []$
		\ForAll{$m in models$}
			\State $m.$\Call{Terminate}{}
			\State \textbf{Append} $allErrors$, $m.errors$
		\EndFor
		\State \Call{Error}{$allErrors$}
	\Else
		\ForAll{$m in models$}
			\State $m.$\Call{Stop}{}
		\EndFor
	\EndIf
	\end{algorithmic}
\end{algorithm}
%\end{center}
%
\todo{make sure this is split correctly across pages}

@@

{\pkg} includes its own implementation of asynchronous communication via each of the communication mechanisms that are supported. While there are existing tools for asynchronous communication using each of these tools, using a tool developed specifically for {\pkg} allows for a more uniform treatment of the different communication mechanisms, greater control over how threads are managed (e.g. killing a thread when messages cannot be interpreted), and a more seamless coordination with other {\pkg} processes (e.g. enforcing dependencies on other model and connection threads).

@@

@@

%---------------------------------------------------------------------------------
% ASYNCHRONOUS MESSAGE MASSING
\subsubsection{Asynchronous Message Passing}\label{SSS:async}
%
The same asynchronous communication strategy is used with each of the communication mechanisms supported by {\pkg}. 
Models do not block on sending messages to 
output channels; the model is free to continue working on its task while 
an \emph{output driver} waits for the message to be routed and received on a separate master process thread. Similarly, 
an \emph{input driver} continuously pings input channels on another thread, moving received messages into a 
buffer queue so that they are ready and waiting for the receiving model 
when it asks for input. These drivers work together to move message along from one model to the next like a conveyor belt. 
As a result, models in complex integration networks are not affected by the rate at which dependent model consume their output and the speedup offered by running the models in parallel is improved. Figure \ref{fig:async} describes the general flow of messages, Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} describe the asynchronous output \& input channel procedures, and Algorithm \ref{alg:conn} describes the procedure followed by input \& output connection drivers.
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/io_drivers.png}
	\caption{Diagram of how messages are passed asynchronously using input/output drivers and an intermediate channel.}
	\label{fig:async}
	\end{center}
	\end{figure}
\fi
%
\begin{enumerate}
	\item A model sends a message in the form of a native data object via a language-specific API to one of the output channels declared in the model YAML. The output channel interface encodes the message and sends it.
	\item An output connection driver (written in Python) runs in a separate thread on the master process, listening to the model output channel. When the model sends a message, the output connection driver checks that the message is in the expected format and then forwards it to an intermediate channel. The intermediate channel is used as a buffer for support on distributed architectures (e.g. if one model is running on a remote machine). In these cases, the intermediate channel will connect to a RMQ broker using security credentials.
	\item An input connection driver (also written in Python) runs in a separate thread, listening to the intermediate channel. When a message is received, it is then forwarded to the input channel of the receiving model as specified in the integration network YAML.
	\item The receiving model receives the message in the form of an analogous native data object. Interface receive calls can be either blocking or non-blocking, but are blocking by default.
\end{enumerate}

% SENDER
\begin{algorithm}[H]
    \caption{Asynchronous communication class for sending messages to an output channel. The procedures {\sc Open, Close, Send, \& IsOpen} are procedures that are implemented by external communication packages and different slightly from one communication mechanism to the next.}
    \label{alg:async_send}
    \begin{algorithmic}[1]
    % SENDER
    \Class{AsyncSender}{$address$}
    	\State $backlog \gets []$
	\State $x \gets$ \Call{Open}{$address$}
	\State \textbf{Run} \Call{BacklogThread}{} on a new thread
	\item[]
	%
	\ClassMethod{BacklogThread}{}
		\While{\Call{IsOpen}{$x$}}
    			\If{$length(backlog) > 0$}
				\State $flag \gets$ \Call{Send}{$x$, $backlog[0]$}
				\If{$flag$}
					\State \textbf{Remove} $backlog[0]$
				\Else
					\State \Call{Close}{$x$}
				\EndIf
			\Else
				\State \textbf{Sleep}
			\EndIf
    		\EndWhile
    	\EndClassMethod
	\item[]
	%
	\ClassMethod{SendBacklog}{$msg$}
		\If{\Call{IsOpen}{$x$}}
			\State \textbf{Append} $backlog$, $msg$
		\EndIf
		\State \textbf{Return} \Call{IsOpen}{$x$}
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Close}{}
		\State \Call{Close}{$x$}
		\State $backlog \gets []$
	\EndClassMethod
	\item[]
    \EndClass
    \end{algorithmic}
\end{algorithm}

% RECEIVER
\begin{algorithm}[H]
    \caption{Asynchronous communication class for receiving messages from an input channel. The procedures {\sc Open, Close, Recv, IsOpen, \& IsMessageWaiting} are procedures that are implemented by external communication packages and different slightly from one communication mechanism to the next.}
    \label{alg:async_recv}
    \begin{algorithmic}[1]
    \item[]
    \Class{AsyncReceiver}{$address$}
    	\State $backlog \gets []$
	\State $x \gets$ \Call{Open}{$address$}
	\State $thread \gets$ \Call{BacklogThread}{} on a new thread
	\item[]
	%
	\ClassMethod{BacklogThread}{}
		\While{\Call{IsOpen}{$x$}}
    			\If{\Call{IsMessageWaiting}{$x$}}
				\State $flag, msg \gets$ \Call{Recv}{$x$}
				\If{$flag$}
					\State \textbf{Append} $backlog$, $msg$
				\Else
					\State \Call{Close}{$x$}
				\EndIf
			\Else
				\State \textbf{Sleep}
			\EndIf
		\EndWhile
	\EndClassMethod
	\item[]
	%
	\ClassMethod{RecvBacklog}{$msg$}
		\If{$length(backlog) > 0$}
			\State $msg \gets backlog[0]$
			\State \textbf{Remove} $backlog[0]$
			\State \textbf{Return} true, $msg$
		\Else
			\State \textbf{Return} \Call{IsOpen}{$x$}, ''
		\EndIf
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Close}{}
		\State \Call{Close}{$x$}
		\State $backlog \gets []$
	\EndClassMethod
	\item[]
    \EndClass
    \end{algorithmic}
\end{algorithm}
%

%
\begin{algorithm}[H]
    \caption{Connection Driver}
    \label{alg:conn}
    \begin{algorithmic}[1]
    \Class{ConnectionDriver}{$inputAddress, outputAddress, yml$}
    	\State $input \gets$ \Call{AsyncReceiver}{$inputAddress$}
    	\State $output \gets$ \Call{AsyncSender}{$outputAddress$}
	\State $thread \gets$ \Call{ConnectionThread}{} on a new thread
	\item[]
	%
	\ClassMethod{ConnectionThread}{}
		\While{true}
			\State $flag, msg \gets$ $input.$\Call{RecvBacklog}{}
			\If{not $flag$} \State \textbf{break} \EndIf
			\If{$msg$}
				\State \textbf{Transform} $msg$ based on units and expected data format
				\State $flag \gets$ $output.$\Call{SendBacklog}{$msg$}
				\If{not $flag$} \State \textbf{break} \EndIf
			\Else
				\State \textbf{Sleep}
			\EndIf
		\EndWhile
		%
		\State $input.$\Call{CloseAsync}{}
		\State $output.$\Call{CloseAsync}{}
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Terminate}{}
		\State $input.$\Call{CloseAsync}{}
		\State $output.$\Call{CloseAsync}{}
	\EndClassMethod
	\item[]
	%
	\ClassMethod{Stop}{}
		\While{\Call{IsOpen}{$input.x$} and \Call{IsOpen}{$output.x$}}
			\State \textbf{Sleep}
		\EndWhile
		\State \Call{Terminate}{}
	\EndClassMethod
	\item[]
    \EndClass
    \end{algorithmic}
\end{algorithm}

\todo{discuss algorithm}

@@

%---------------------------------------------------------------------------------
% SENDING/RECEIVING LARGE MESSAGES
\subsubsection{Sending/Receiving Large Messages}\label{SSS:large}
%
All of the communication tools leveraged by {\pkg} have intrinsic limits 
on the allowed size for a single message. Some of these limits can be quite large 
($2^{20}$ bytes for ZMQ and RMQ), while others are very limiting 
(2048 bytes on Mac OS X for IPC queues). Although messages consisting of a few scalars are unlikely to 
exceed these limits, biological inputs and outputs are often much more complex. 
For example, structural data represented as a 3D mesh can easily exceed these 
limits. To handle messages that are larger than the limit of the communication 
mechanism being used, {\pkg} splits the message up into multiple smaller 
messages. In addition, for large messages, {\pkg} creates new, temporary 
communication channels that are used exclusively for a single message and 
then destroyed. The address associated with the temporary channel is sent in 
header information as a message on the main channel along with metadata about 
the message that will be sent through the temporary channel like size and data type. 
Temporary channels are used for large messages to prevent mistakenly combining 
the pieces from two different large messages that were received at the same time 
such as in the case that a model is receiving input from two different models 
working in parallel.

%\begin{algorithm}[H]
%    \caption{Large Messages}
%    \label{alg:large}
%    \begin{algorithmic}[1] 
%    \end{algorithmic}
%\end{algorithm}

\todo{algorithm}

@@

In addition to basic input and output, each interface also provides access 
to more complex data types and communication patterns.

@@

It should also be noted that it is possible to pass multiple variables as a single input/output. However, unless the variables will \emph{always} be coupled in the model, it is advised that every variable be specified separately for clarity and to allow flexibility in the way other models can integrate with it. In addition, if brevity is a concern because a model has a large number of input variables, there are features currently under active development (see \S\ref{SS:datagg}) that will allow fields to be aggregated within connection YAMLs and simplify the send and receive calls in such cases.

@@

Integration via {\pkg} requires minimal modification of the existing model source code yet greatly increases the model's potential by allowing it to be reused in combination with other models. {\pkg} has already been used to integrate several plant models within the Crops in Silico organization \citep{Marshall-Colon2017} with ongoing work to add additional models to the integration network. Figure \ref{fig:network} shows the progress so far. 
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/CiS-Languages.png}
	\caption{Progress towards full Crops in Silico integration network. Each square represents a model and the color of the square indicates the language in which the model is written. Arrows between models represent connections being made between the models using {\pkg}. Solid lines are connections that have already been made, dashed lines are in-progress connections, and dotted lines are planned connections.}
	\label{fig:network}
	\end{center}
	\end{figure}
\fi
%
Each square is a model, color indicates the programming language that the model is written in, and lines between the models represent connections where information is exchanged. Solid lines are connections that have already been implemented, dashed lines are connections currently being implemented, and dotted lines are connections planned for the future. When complete, the integration network will include 9 models of molecular, genetic, cellular, organ, and plant level processes in 4 different programming languages, produced by $>10$ researchers at 4 different institutions.

>>>>

\pdfoutput=1
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage[noend]{algpseudocode}
\usepackage{caption}
\algblockdefx[Try]{Try}{EndTry}%
	{\textbf{try}}%
	{\textbf{end try}}
\algcblockdefx[TryExcept]{Try}{Except}{EndTry}%
	{\textbf{except}}%
	{\textbf{end try}}
\algblockdefx[class]{Class}{EndClass}%
	[2]{\textbf{class} {\sc #1}(#2)}%
	{\textbf{end class}}
\algblockdefx[classmethod]{ClassMethod}{EndClassMethod}%
	[2]{\textbf{method} {\sc #1}(#2)}%
	{\textbf{end method}}
\usepackage[toc,page]{appendix}

@@

%---------------------------------------------------------------------------------
% ABSTRACT (145 words)
\begin{abstract}
Thousands of computational models have been created within the plant biology community and broader scientific communities in the past two decades that have the potential to be combined into complex integration networks capable of capturing more complex biological processes than possible with isolated models. However, the technological barriers introduced by differences in language and data formats has slowed this progress. We present \href{https://github.com/cropsinsilico/cis_interface}{\pkg}, a Python package for running integration networks with connections between models across languages and scales. {\pkg} coordinates parallel execution of models in Python, C, C++, and Matlab on Linux, Mac OS, and Windows operating systems, and handles communication in a number of data formats common to computational plant modeling. {\pkg} is designed to be user-friendly and can be accessed at {\tt https://github.com/cropsinsilico/cis\_interface}. Although originally developed for plant models, {\pkg} can be used to connect computational models from any domain.
\end{abstract}

@@

%---------------------------------------------------------------------------------
% DOMAIN SPECIFIC LANGUAGE
\subsection{Domain Specific Languages (DSLs)}\label{SS:dsl}
%
One solution is to agree upon a language. The questions of orchestration, communication, and translation become trivial when models are all written in the same language. Beyond selecting a single programming language, many communities have  developed dedicated domain specific languages (DSLs) for model representation. For example, the Systems Biology Markup Language \citep[SBML][]{Hucka2003} is an XML-based format for representing models of biological processes. Scientists can compose their models using SBML markup and then run their model using software designed to parse SBML files. Other DSLs for biological models include LPy \citep{Boudon2012}, CellML \citep{Cuellar2003}, FLAME \citep{Coakley2012}, and BioPAX \citep{Demir2010}. Dedicated model DSLs improve the reusability of models written in the DSL and have the advantage of often implicitly handling data transformation. However, they provide no help for existing models that are not written in the DSL or those that cannot be expressed within the constraints of the DSL. Furthermore, learning a new DSL, in addition to a programming language, can be daunting.

@@

There are also domain specific frameworks for connecting biological models. For example, OpenAlea \citep{Pradal2015} allows users to compose networks from different components including plant models, analysis tools, and visualization tools. Domain specific frameworks are more intuitive and ultimately more flexible than DSLs in the types of operations they allow models to perform since they have access to the full power of a programming language. However, they are stricter in several respects: 1) the model must be written in (or exposable to), the language of the framework and 2) the model must be written in a way that is aware of the framework and/or the format of models with which it will interact. Like DSLs, frameworks also often provided limited support for models written without the framework in mind.

@@

{\pkg} was designed to transform models into building blocks that can be easily combined with other blocks to build complex structures, thus promoting model reuse and collaboration between scientists with varying degrees of overlapping expertise. {\pkg} does so by addressing the questions from \S\ref{S:background} while being:

@@

{\pkg} is built upon several well-established open-source software packages(see below) along with new tools developed explicitly for {\pkg} including utilities for running and monitoring models written in other languages from Python, dynamically creating and managing communication networks, data type conversions between languages, and asynchronous message passing with variable underlying communication mechanisms. {\pkg} currently support models written in Python, Matlab, C, and C++ with additional Domain Specific Language (DSL) support for LPy models \citep{Boudon2012}. Support for additional languages is planned for future development (See \S\ref{SS:langsupport}).

%---------------------------------------------------------------------------------
% ORCHESTRATION
\subsection{Orchestration}\label{SS:orchestration}
%
{\pkg} is executed via a command line interface (CLI), {\pkgrun} with specification files (\S\ref{SSS:yaml}) as input. Based on the information contained in the specification files, {\pkg} dynamically establishes a network of asynchronous communication channels (\S\ref{SS:communication}), and launches the models on new processes (\S\ref{SS:model_execution}). Although written in Python, {\pkg} was written such that users need not have any knowledge of the Python language and can interact with {\pkg} solely through the CLI.
%At execution, {\pkg} launches models on new processes based on information contained in a YAML specification file (See \S\ref{SSS:yaml} below) and establishes

@@

%---------------------------------------------------------------------------------
% MODEL EXECUTION
\subsubsection{Model Execution}\label{SS:model_execution}
%
{\pkg} launches each model in an integration on its own process, allowing models to complete independent operations in parallel and complete tasks more quickly. For example, the root and shoot growth models from \S\ref{S:intro} require 2.42\,s and 4.06\,s respectively to run for 100 time steps in isolation. If these two models were manually integrated in serial via the command line, the integration would require a total of 6.48\,s. However, the same integration requires only 4.52\,s using {\pkg}, a speed-up of 1.43. The exact speedup provided by parallel integration using {\pkg} depends on how independent the models are and how equally distributed the work is between the models. If the models are fully independent, the speedup is limited by the time required to execute the slowest model. In the example, this would be the shoot growth model which required almost twice the time to execute as the root model. If the models are dependent on one-another, the models will have to wait for messages to be passed and the speedup will not be as great (see \S\ref{SS:speedup} for additional discussion of parallelism and the speedups achievable using {\pkg}). \todo{Fix these numbers for most recent version of the code}.

While every model is executed on a new process, how the model is handled depends 
on the language it is written in. {\pkg} has a dedicated driver for each of the supported languages with utilities that allow {\pkg} to launch and monitor executables written in that language from Python. Models written in 
interpreted languages (Python and Matlab) are executed on the command line 
with the interpreter. In the case of Matlab, where a significant amount of time is required to start 
the Matlab interpreter (See \S\ref{SS:results_language}), Matlab shared engines are used 
to execute Matlab models. Matlab shared engines are Matlab instances that other processes can 
submit Matlab code to for execution. While shared engines also require the same amount of time 
to start as a standard instance of the Matlab interpreter, they can be started in advance and then 
reused.

@@

The algorithms for a generic model driver and the master thread can be found in Algorithms \ref{alg:model} \& \ref{alg:master} respectively in \A\ref{A:algorithms}.

@@

{\pkg} includes its own implementation of asynchronous communication via each of the communication mechanisms that are supported. While there are existing tools for asynchronous communication using each of these communication mechanisms, using a tool developed specifically for {\pkg} allows for a more uniform treatment of the different communication mechanisms, greater control over how threads are managed (e.g. killing a thread when messages cannot be interpreted), and a more seamless coordination with other {\pkg} processes (e.g. enforcing dependencies on other model and connection threads).

@@

In addition to using the default ZMQ libraries, {\pkg} also includes supporting routines for allowing received messages to be confirmed. Because ZMQ is broker-less, a socket has no way of knowing if the message it sent was successfully received. This lack of confirmation makes it harder to determine if there is was error somewhere along the network. {\pkg} overcomes this by generating two ZMQ sockets for every connection: one for sending the messages and one for confirming them. Each time a message is received, the receiving socket confirmation thread will send a confirmation including a unique ID for the received message and then wait for a reply to that confirmation. The sending socket confirmation thread will continuously check the confirmation sockets for messages indicating that sent messages were received. Once a confirmation message is received, it will send a reply to the receiving confirmation socket and record the ID of the message that was confirmed. This handshake operation ensures that all messages are accounted for so {\pkg} knows if a message is lost and where it was lost.

@@

%---------------------------------------------------------------------------------
% ASYNCHRONOUS MESSAGE MASSING
\subsubsection{Asynchronous Message Passing}\label{SSS:async}
%
The same asynchronous communication strategy is used with each of the communication mechanisms supported by {\pkg}. 
Models do not block on sending messages to 
output channels; the model is free to continue working on its task while 
an \emph{output driver} waits for the message to be routed and received on a separate master process thread. Similarly, 
an \emph{input driver} continuously pings input channels on another thread, moving received messages into a 
buffer queue so that they are ready and waiting for the receiving model 
when it asks for input. These drivers work together to move message along from one model to the next like a conveyor belt. 
As a result, models in complex integration networks are not affected by the rate at which dependent model consume their output and the speedup offered by running the models in parallel is improved. Figure \ref{fig:async} describes the general flow of messages, Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} in \A\ref{A:algorithms} describe the asynchronous output \& input channel procedures, and Algorithm \ref{alg:conn} describes the procedure followed by input \& output connection drivers.
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/io_drivers.png}
	\caption{Diagram of how messages are passed asynchronously using input/output drivers and an intermediate channel.}
	\label{fig:async}
	\end{center}
	\end{figure}
\fi
%
\begin{enumerate}
	\item A model sends a message in the form of a native data object via a language-specific API to one of the output channels declared in the model YAML. The output channel interface encodes the message and sends it.
	\item An output connection driver (written in Python) runs in a separate thread on the master process, listening to the model output channel. When the model sends a message, the output connection driver checks that the message is in the expected format and then forwards it to an intermediate channel. The intermediate channel is used as a buffer for support on distributed architectures (e.g. if one model is running on a remote machine). In these cases, the intermediate channel will connect to a RMQ broker using security credentials.
	\item An input connection driver (also written in Python) runs in a separate thread, listening to the intermediate channel. When a message is received, it is then forwarded to the input channel of the receiving model as specified in the integration network YAML.
	\item The receiving model receives the message in the form of an analogous native data object. Interface receive calls can be either blocking or non-blocking, but are blocking by default.
\end{enumerate}

@@

%---------------------------------------------------------------------------------
% SENDING/RECEIVING LARGE MESSAGES
\subsubsection{Sending/Receiving Large Messages}\label{SSS:large}
%
All of the communication tools leveraged by {\pkg} have intrinsic limits 
on the allowed size for a single message. Some of these limits can be quite large 
($2^{20}$ bytes for ZMQ and RMQ), while others are very limiting 
(2048 bytes on Mac OS X for IPC queues). Although messages consisting of a few scalars are unlikely to 
exceed these limits, biological inputs and outputs are often much more complex. 
For example, structural data represented as a 3D mesh can easily exceed these 
limits. To handle messages that are larger than the limit of the communication 
mechanism being used, {\pkg} splits the message up into multiple smaller 
messages. In addition, for large messages, {\pkg} creates new, temporary 
communication channels that are used exclusively for a single message and 
then destroyed. The address associated with the temporary channel is sent in 
header information as a message on the main channel along with metadata about 
the message that will be sent through the temporary channel like size and data type. 
Temporary channels are used for large messages to prevent mistakenly combining 
the pieces from two different large messages that were received at the same time 
such as in the case that a model is receiving input from two different models 
working in parallel. The procedures for creating and using temporary channels for sending large messages can be seen in the {\sc Send} and {\sc Recv} methods from Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} respectively in \A\ref{A:algorithms}.

@@

In addition to basic input and output, each interface also provides access 
to more complex data types and communication patterns. Pseudocode describing the asynchronous communication classes and functions underlying the interfaces is provided by Algorithms \ref{alg:async_send} \& \ref{alg:async_recv} in \A\ref{A:algorithms}.

@@

It should also be noted that it is possible to pass multiple variables as a single input/output. However, unless the variables will \emph{always} be coupled in the model, it is advised that every variable be specified separately for clarity and to allow flexibility in the way other models can integrate with it. In addition, if brevity is a concern because a model has a large number of input variables, there are features currently under active development (see \S\ref{SS:dataagg}) that will allow fields to be aggregated within connection YAMLs and simplify the send and receive calls in such cases.

@@

{\pkg} is designed to make models as reusable as possible. Once integrated via {\pkg}, models can be combined with any other models that also have a {\pkg} wrapper and YAML specification file with minimal, if any, modification. This "plug-and-play" approach to models encourages model writers to make their codes public in a reusable form, generating citations and reuse of their work, while also growing a community of models and model writers capable of probing areas otherwise unreachable. 
%Generic workflow, parallelization, and communication tools can be used to expose/connect models, but without a standard for making connections, every new integration will require a great deal of hands on attention and new code. Domain specific frameworks like OpenAlea \citep{Pradal2008} are better for improving the reusability for some models in that they can impose a standard for how models interact with each other, however none of the existing frameworks offer the same flexibility as {\pkg} in terms of languages with native support and model form.
%Integration via {\pkg} requires minimal modification of the existing model source code yet greatly increases the model's potential by allowing it to be reused in combination with other models. 
%
{\pkg} has already been used to integrate several plant models within the Crops in Silico organization \citep{Marshall-Colon2017} with ongoing work to add additional models to the integration network. Figure \ref{fig:network} shows the progress so far. 
%
\ifinclfig
 	\begin{figure}[htbp]
	\begin{center}
	\includegraphics[width=\columnwidth,keepaspectratio]{./images/CiS-Languages.png}
	\caption{Progress towards full Crops in Silico integration network. Each square represents a model and the color of the square indicates the language in which the model is written. Arrows between models represent connections being made between the models using {\pkg}. Solid lines are connections that have already been made, dashed lines are in-progress connections, and dotted lines are planned connections.}
	\label{fig:network}
	\end{center}
	\end{figure}
\fi
%
Each square is a model, color indicates the programming language that the model is written in, and lines between the models represent connections where information is exchanged. Solid lines are connections that have already been implemented, dashed lines are connections currently being implemented, and dotted lines are connections planned for the future. When complete, the integration network will include 9 models of molecular, genetic, cellular, organ, and plant level processes in 4 different programming languages, produced by $>10$ researchers at 4 different institutions.
